---
author: Ben Scott
date: "`r Sys.Date()`"
title: "Theil-Sen Trend Analysis Using Environment Canterbury Data"
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
execute:
  warning: false
  message: false
params:
  from_date: "01/01/2015"
  to_date: "31/12/2024"
  output_dir: "outputs"
---

```{r}
#| label: setup

knitr::opts_chunk$set(echo = FALSE)

required_packages <- c(
  "dplyr",
  "knitr",
  "lubridate",
  "openair",
  "readr",
  "tidyr"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the required analysis packages: ",
    paste(missing_packages, collapse = ", "),
    call. = FALSE
  )
}

library(dplyr, warn.conflicts = FALSE)
library(lubridate, warn.conflicts = FALSE)
library(openair, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
```

## Introduction

This report retrieves daily air-quality data from the Environment Canterbury
(ECan) data portal, prepares PM10 and PM2.5 data for OpenAir, checks coverage,
and calculates Theil-Sen trends.

The dates, included sites, and exclusions are analysis decisions. Review them
before each annual run.

## Data retrieval

`get_daily_all_stations()` retrieves daily observations for every station
available from ECan for the requested period.

```{r}
#| label: get-data

df_final_long <- ecanairquality::get_daily_all_stations(
  from_date = format(dmy(params$from_date), "%d/%m/%Y"),
  to_date = format(dmy(params$to_date), "%d/%m/%Y")
)
```

The returned data is tidy and long-format, with one row per station, date, and
measurement.

## Data cleaning and analysis decisions

OpenAir expects a `date` column and pollutant measurements in separate numeric
columns. The following transformation keeps only PM10 and PM2.5, creates the
OpenAir station column, and adds coverage fields useful for review.

```{r}
#| label: data-cleaning

pollutant_names <- c("PM10 (ug/m3)", "PM25 (ug/m3)")

df_daily_pm <- df_final_long |>
  filter(name %in% pollutant_names) |>
  pivot_wider(
    names_from = name,
    values_from = value
  ) |>
  rename(
    date = DateTime,
    site = StationName,
    PM10 = `PM10 (ug/m3)`,
    PM25 = `PM25 (ug/m3)`
  ) |>
  mutate(
    date = ymd(date, tz = "Pacific/Auckland"),
    month = month(date),
    year = year(date),
    days_per_month = days_in_month(date)
  )
```

The default exclusions below reflect the current ECan trend-analysis
decisions. Update them when monitoring sites move, close, or change purpose.

```{r}
#| label: apply-analysis-decisions

excluded_sites <- c("Washdyke Flat Road", "Waimate Stadium", "St Albans EP")

df_daily_pm_clean <- df_daily_pm |>
  filter(!site %in% excluded_sites)
```

## Coverage checks

Use summary plots to inspect missing periods and confirm that the selected
sites and dates are suitable before calculating trends.

```{r}
#| label: pm10-summary-plot
#| fig-width: 20
#| fig-height: 30

openair::summaryPlot(
  df_daily_pm_clean |> select(date, site, PM10),
  pollutant = "PM10",
  type = "site",
  fontsize = 20
)
```

```{r}
#| label: pm25-summary-plot
#| fig-width: 20
#| fig-height: 30

openair::summaryPlot(
  df_daily_pm_clean |> select(date, site, PM25),
  pollutant = "PM25",
  type = "site",
  fontsize = 20
)
```

The plots should be reviewed for long gaps before accepting the trend output.
In particular, investigate gaps longer than five months and apply the
relevant period or site decision before proceeding.

## Theil-Sen analysis

Theil-Sen estimates are calculated separately for PM10 and PM2.5, with
seasonality removed and a 75% data threshold.

```{r}
#| label: trend-analysis-pm10
#| fig-width: 20
#| fig-height: 20

openair::TheilSen(
  df_daily_pm_clean,
  ylab = "PM10 (ug/m3)",
  pollutant = "PM10",
  deseason = TRUE,
  date.format = "%Y",
  type = "site",
  data.thresh = 75,
  fontsize = 30
)
```

```{r}
#| label: trend-analysis-pm25
#| fig-width: 20
#| fig-height: 20

openair::TheilSen(
  df_daily_pm_clean,
  ylab = "PM2.5 (ug/m3)",
  pollutant = "PM25",
  deseason = TRUE,
  date.format = "%Y",
  type = "site",
  data.thresh = 75,
  fontsize = 30
)
```

## Data export

Save the exact retrieved long-format data and the cleaned analysis data so
results can be audited or reproduced without another ECan request.

```{r}
#| label: data-export

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)

date_stamp <- Sys.Date()

readr::write_csv(
  df_final_long,
  file.path(params$output_dir, paste0("ecan_daily_long_", date_stamp, ".csv"))
)

readr::write_csv(
  df_daily_pm_clean,
  file.path(params$output_dir, paste0("ecan_pm_trend_input_", date_stamp, ".csv"))
)
```

The generated files record the source data and the analysis-ready data used by
OpenAir. Keep them with the rendered report when results need to be compared
between runs.
